A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-11-08 Draft14.tex
Update Draft14.tex

\paragraph{Problem Setting}
This paper is concerned with the class of constrained optimization problems
\begin{align}\label{eq:main-problem}
\min_{x \in \X} \; f(x),
\end{align}
where $\X \subseteq \R^d$ is a closed convex set and $f \colon \X \to \R$ is a continuously differentiable convex function. 
% We primarily focus on the case $\X = \R^d$ unless explicitly noted otherwise. 
%We also assume $f$ has a unique minimizer, $x^* \in \X$, satisfying the optimality condition $\nabla f(x^*) = 0$. 
We use the standard Euclidean norm $\|x\| = \langle x,x \rangle^{1/2}$ throughout.
We consider the general non-Euclidean setting in which the space $\X$ is endowed with a distance-generating function $h \colon \X \to \R$ that is convex and essentially smooth (i.e., $h$ is continuously differentiable in $\X$, and $\|\nabla h(x)\|_* \to \infty$ as $\|x\| \to \infty$). The function $h$ can be used to define a measure of distance in $\X$ via its Bregman divergence:
\begin{align*}
D_h(y,x) = h(y) - h(x) - \langle \nabla h(x), y-x \rangle,
\end{align*}
which is nonnegative since $h$ is convex. 
The \emph{Euclidean setting} is obtained when $h(x) = \frac{1}{2} \|x\|^2$.

@@

The family of dynamics that obey the ideal scaling constraints~\eqref{Eq:IdeSca} is invariant under time reparametrizations.  Such time-invariance properties provide more flexibility in algorithm design and analysis, as we will see throughout this paper.

@@

\paragraph{General Accelerated Mirror Descent} In Wibisono {\em et al.}~\cite{Acceleration}, the following Lyapunov function:
\begin{align}\label{Eq:E}
\E_t \,=\, e^{\beta_t} (f(X_t)-f(x^*)) + D_h\left(x^*, \, X_t + e^{-\alpha_t} \dot X_t\right),
\end{align}
was introduced to analyze the flow that forms the basis for accelerated mirror descent.
%  Notice, assuming we start from rest $\dot X_0 = 0$, the  Lyapunov property allows us to conclude,
 %\begin{equation*}
 %f(X_t) - f(x^\ast)\leq \frac{e^{\beta_0} (f(X_0) - f(x^\ast) )+ D_h(x^\ast, X_0)}{e^{\beta_t}}.
 %\end{equation*}
In Section~\ref{Sec:MomLyap}, we show how this Lyapunov function emerges from a general framework for analyzing momentum dynamics; furthermore, we show how to discretize the Lyapunov function and thereby obtain a tool which can be used to analyze a variety of algorithms that are discretizations of momentum dynamics.
%MJ:  I don't view this footnote as necessary; that derivation was in supplementary information \footnote{The same derivation of the Lyapunov function was added to~\cite[App. C]{Acceleration}}
%
%
%
%ACCC GRAD STRONG
\paragraph{Accelerated Gradient Descent (Strong Convexity)}
In this paper, we show that the following function 
   \begin{equation}\label{eq:general-sc-lyap}
  \E_t =  e^{\beta_t} \left(f(X_t) - f(x^\ast) + \mu D_h\Big(x^\ast ,X_t + e^{-\alpha_t} \dot X_t\Big)\right),
 \end{equation}
is a Lyapunov function for a family of momentum dynamics when $f$ is $\mu$-strongly convex ($D_f(y,x) \geq \mu D_h(y,x)$) and $e^{\alpha_t} = \dot \beta_t$.  In Section~\ref{Sec:StrongConv}, we derive this Lyapunov function and show how it can be used to demonstrate a linear rate of convergence for the family of accelerated gradient descent dynamics.  (For simplicity, we focus on the setting in which $h = \frac{1}{2}\|\cdot\|^2$ is Euclidean.) We also show that a particular discretization of the family of dynamics provides an algorithm for which a commensurate discretization of~\eqref{eq:general-sc-lyap} is a Lyapunov function.

>>>>

\paragraph{Problem setting}
This paper is concerned with the class of constrained optimization problems
\begin{align}\label{eq:main-problem}
\min_{x \in \X} \; f(x),
\end{align}
where $\X \subseteq \R^d$ is a closed convex set and $f \colon \X \to \R$ is a continuously differentiable convex function. 
% We primarily focus on the case $\X = \R^d$ unless explicitly noted otherwise. 
%We also assume $f$ has a unique minimizer, $x^* \in \X$, satisfying the optimality condition $\nabla f(x^*) = 0$. 
We use the standard Euclidean norm $\|x\| = \langle x,x \rangle^{1/2}$ throughout.
We consider the general non-Euclidean setting in which the space $\X$ is endowed with a distance-generating function $h \colon \X \to \R$ that is convex and essentially smooth (i.e., $h$ is continuously differentiable in $\X$, and $\|\nabla h(x)\|_* \to \infty$ as $\|x\| \to \infty$). The function $h$ can be used to define a measure of distance in $\X$ via its Bregman divergence:
\begin{align*}
D_h(y,x) = h(y) - h(x) - \langle \nabla h(x), y-x \rangle,
\end{align*}
which is nonnegative since $h$ is convex. 
The \emph{Euclidean setting} is obtained when $h(x) = \frac{1}{2} \|x\|^2$.

@@

The family of dynamics that obey the ideal scaling constraints~\eqref{Eq:IdeSca} is invariant under time reparameterizations.  Such time-invariance properties provide more flexibility in algorithm design and analysis, as we will see throughout this paper.

@@

\paragraph{General accelerated mirror descent} In Wibisono {\em et al.}~\cite{Acceleration}, the following Lyapunov function:
\begin{align}\label{Eq:E}
\E_t \,=\, e^{\beta_t} (f(X_t)-f(x^*)) + D_h\left(x^*, \, X_t + e^{-\alpha_t} \dot X_t\right),
\end{align}
was introduced to analyze the flow that forms the basis for accelerated mirror descent.
%  Notice, assuming we start from rest $\dot X_0 = 0$, the  Lyapunov property allows us to conclude,
 %\begin{equation*}
 %f(X_t) - f(x^\ast)\leq \frac{e^{\beta_0} (f(X_0) - f(x^\ast) )+ D_h(x^\ast, X_0)}{e^{\beta_t}}.
 %\end{equation*}
In Section~\ref{Sec:MomLyap}, we show how this Lyapunov function emerges from a general framework for analyzing momentum dynamics; furthermore, we show how to discretize the Lyapunov function and thereby obtain a tool which can be used to analyze a variety of algorithms that are discretizations of momentum dynamics.
%MJ:  I don't view this footnote as necessary; that derivation was in supplementary information \footnote{The same derivation of the Lyapunov function was added to~\cite[App. C]{Acceleration}}
%
%
%
%ACCC GRAD STRONG
\paragraph{Accelerated Gradient Descent (Strong Convexity)}
In this paper, we show that the following function 
   \begin{equation}\label{eq:general-sc-lyap}
  \E_t =  e^{\beta_t} \left(f(X_t) - f(x^\ast) + \mu D_h\Big(x^\ast ,X_t + e^{-\alpha_t} \dot X_t\Big)\right),
 \end{equation}
is a Lyapunov function for a family of momentum dynamics when $f$ is $\mu$-strongly convex ($D_f(y,x) \geq \mu D_h(y,x)$) and $e^{\alpha_t} = \dot \beta_t$.  In Section~\ref{Sec:StrongConv}, we derive this Lyapunov function and show how it can be used to demonstrate a linear rate of convergence for the family of accelerated gradient descent dynamics.  (For simplicity, we focus on the setting in which $h = \frac{1}{2}\|\cdot\|^2$ is Euclidean.) We also show that a particular discretization of the family of dynamics provides an algorithm for which a commensurate discretization of~\eqref{eq:general-sc-lyap} is a Lyapunov function.
